Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-07-09 POPL/paper.tex
syntax section minus a couple cases

Omission of the structural inference rules of a deductive system, such as in
linear logic \cite{GIRARD19871}, offers precise control over the number of times
a value may be used in a derivation. This control enables
\emph{resource-sensitive} reasoning, where we may treat a resource as
\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
as we may treat characters of a string finite resources that are consumed at
parse-time. One may then envision a linear type system where the types comprise
formal grammars generated over some alphabet $\Sigma$, and the type constructors
correspond precisely to inductive constructions on grammars --- such as
conjunction, disjunction, and concatenation.

@@

\steven{Maybe there is more to say here, or perhaps later, that the only thing
  to gain from a non-dependent non-linear fragment are the things that are
  definiable via $G$. Except that these are not interesting, as they are
  necessarily empty}

@@

In this paper, we instantiate an LNL theory with dependent types as a deductive
setting to conduct parsing and formal grammar theory. In this system, linear
types encode formal grammars built over an alphabet/(set of base types)
$\Sigma$. The non-linear types are just MLTT/math defs. The syntax is in the
following subsections

\steven{Rephrase the above paragraph to summarize the intro of this section, as
  well as preview the syntax that's to come in the following subsections.}

\steven{Everything below here hasn't really been touched}

\subsection{Syntax}

\steven{Next paragraph is paper scaffolding for what I want to say and is total
  word salad rn} In this paper, we instantiate an LNL theory to axiomatize the
notion of a Chomsky category, and because $\Set^{\String}$ forms a Chomsky
category, using the type theory corresponds to interesting formal grammar
proofs. Internalization of previous results + use to build
correct-by-construction parsers.

\steven{Expand here on how linearity has an intuitive parsing intrerpration, however working only in a simply typed system is not sufficient. i.e. dependence offers the right view on indexed grammars. }

The structrual judgements are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
\cref{fig:inttyping}. There are all just as they appear in
\cite{krishnaswami_integrating_2015}.

Our contribution is the addition of base types drawn from the alphabet $\Sigma$
along with the omission of the exchange rule for linear typing judgements.

\steven{TODO need to update syntax in these rules}

@@

    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \\
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgmental equality}
  \label{fig:jdgeq}
\end{figure}
\subsection{Syntax}
\label{subsec:syntax}

\paragraph{Non-linear Types}

The non-linear fragment of the theory
corresponds to usual Martin-L\"of type theory. That is, the non-linear
contexts obey all of the usual structural rules --- contraction, weakening, and
exchange --- so we may use their contents as many times as we'd like. For the application of parsing, we think of these as primarily as indicies terms for indexed grammars. That is, they are as useful for their dependence. We may think of them as brining into scope definitions from ordinary mathematics to be used for indexing

\paragraph{Linear Types}

The linear typing judgment in our syntax takes on the following schematic form $\Gamma ; \Delta \vdash a : A$. First, $A$ represents
a \emph{linear type} in our syntax. The intended semantics of these linear types
are formal grammars. That is, the linear typing system is designed to syntactically
reflect the behavior of formal grammars. For this reason, we may often
interchangeably use the terms ``linear type'' and ``grammar''.

The term $a$ is an inhabitant of type $A$, which we want to think of as a parse tree of
the grammar $A$. The core idea of this entire paper follows precisely from this
single correspondence: grammars are types, and the inhabitants of these types
are parse trees for the grammars.

$\Gamma$ reprsents a non-linear context, while $\Delta$ represents a \emph{linear context}. These linear
contexts
behave substructurally --- in particular they have no contraction, weakening, or exchange
rule. This means that everything within $\Delta$ must be used \emph{exactly
  once} and \emph{in order of occurrence}. Thus, we can think of the linear
contexts as an ordered list of limited
resources. Once a resource is consumed, we cannot make reference to it again.
We may then
think of the variables in a linear context akin to building blocks for
constructing patterns over strings.

In the realm of parsing, our substructural
restrictions make it so that the linear contexts behave quite similarly to
formal grammars. For instance, linear context extension mirrors concatenation of
grammars which may be seen via the introduction rule for $\otimes$,

\[
  \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta , \Delta' \vdash e \otimes e' : A \otimes B}
\]

Just as two grammars may be concatenated, if we can build an $A$ with the
contents of $\Delta$, and a $B$ with the contents of $\Delta'$, then we we may
concatenate $\Delta$ and $\Delta'$ to build an inhabitant $A \otimes B$.

This is perhaps better illustrated through a concrete example. Consider the
regular expression $A = (a \oplus b) \otimes (b \oplus c)^{*}$ and the string
$w = abc$. We then have the following derivation that $w$ matches $A$,

\begin{equation}
  \label{eq:samplederiv}
  \inferrule
  {
    w_{1} : a \vdash \mathsf{inl}(w_{1}) : a \oplus b \\
    \inferrule
    {
      w_{2} : b \vdash \mathsf{inl}(w_{2}) : b \oplus c \\
      w_{3} : c \vdash \mathsf{inr}(w_{3}) : b \oplus c
    }
    {
      w_{2} : b, w_{3} :c \vdash \mathsf{cons}(w_{2},w_{3}) : (b \oplus c)^{*}
    }
  }
  {w_{1} : a, w_{2} : b, w_{3} : c \vdash \mathsf{inl}(w_{1}) \otimes \mathsf{cons}(w_{2}, w_{3}) : (a \oplus b) \otimes (b \oplus c)^{*} }
\end{equation}

When parsing $w$ into $A$ by hand, we perform the same process as reflected in
the derivation \cref{eq:samplederiv}. That is, we split the string $w$ up into pieces, one of which matches
$a \oplus b$, and and another that matches $(b \oplus c)^{*}$. When further
parsing this Kleene star bit, we break it up into smaller pieces each of which
matches the patterns $b \oplus c$ and then we combine them all into one parse of
the Kleene star $(b \oplus c)^{*}$. Note that this exact process is echoed in
the above proof tree.

\subsubsection{Dependence of the Linear on the Non-Linear}

We may now however ask, ``what are the $\Gamma$'s doing in the $\otimes$ introduction rule?''
Each linear context $\Delta$ is dependent on a non-linear context $\Gamma$, so
we may use variables drawn from $\Gamma$ when defining the types in $\Delta$.

In the grammar setting, this dependence lets
us use variables from regular dependent type theory to
keep track of data without \emph{consuming any resources}, so to speak. That is,
using a piece of a linear context consumes it, so we cannot linearly encode anything that
needs to be referenced more than once.

We will see a concrete instance of this dependence in \cref{subsubsec:pda}. In
that case, we encode the traces of a pushdown automaton as a grammar. When doing so,
we make use of our non-linear types to define a type of stacks. The fact that
our encoding of stacks is non-linear reflects
that a pushdown automaton may consume a character from the input string without
affecting the state of the stack.

\subsubsection{Other Type Constructors}
\label{subsubsec:other}

\paragraph{Brzozowksi Derivative}
In the 1950s Janusz Brzozowksi developed a notion of derivative for formal
languages \cite{brzozowskiDerivativesRegularExpressions1964}. The derivative of language $L$ by string $w$ is given as,

\[
  w^{-1}L = \{ w' \in \String : w \otimes w' \in L \}
\]

That is, the derivative of $L$ with respect to $w$ are precisely the strings
that complete to an $L$-parse when prefixed by a $w$.

Brzozowski initially gave an accounting of this operation for
generalized regular expressions, but later work demonstrates that the same
construction can be generalized to context free grammars \cite{mightParsingDerivativesFunctional2011}.

The ``magic wand'' from linear logic takes on the role of Brzozowski's
derivative in our syntax.

\begin{equation}
  \label{eq:linfun}
\inferrule{\Gamma ; a : A, \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A \lto B}
\end{equation}

We read this as saying that $\Delta$ constructs a $A \lto B$-parse when
prepending $\Delta$ with an $A$-parse completes to a $B$ parse. In this sense,
the type $A \lto B$ behaves like a linear function type. The only distinction
between $A \lto B$ and an intuitionistic function type is that linear functions
consume their input.

Note that the introduction rule presented in \cref{eq:linfun} does indeed
generalize Brzozowski's initial presentation of the derivative. First, to match
his initial presentation
let $A$ be the grammar accepting a single string $w$, which we may write as the
discrete grammar $\lceil w \rceil$.
Additionally, our construction lifts from a language-theoretic construction
to a grammar point-of-view.
In this sense, we carry around not just the data of what strings belong to
$\lceil w \rceil \lto B$ but also are endowed the data of a function that
constructs parse trees of $B$ from pairs of $\lceil w \rceil$-parses and
$(\lceil w \rceil \lto B)$-parses.

\paragraph{Top and Bottom}
We include linear versions of the unit and empty types. The
grammar $\top$ matches every input string, while the grammar
$\bot$ represents falsity and matches no strings. These two
are often used together to give syntactic shorthand to
grammars that encode propositions.

For instance, we may encode the proposition
$\mathsf{isVowel}$ as,

\[
  \mathsf{isVowel}(w) :=
    \begin{cases}
      \top, \quad w \in \{a, e, i, o, u\} \\
      \bot, \quad \text{otherwise}
    \end{cases}
\]

\paragraph{Fixed Point Operator}
We include a least fixed point operator on types, $\mu$, which is the means by
which we define recursive grammars.

@@

% \todoin{with citation, talk somewhere about how CFGs are just regular + fixed points}
\max{this is a good point to give an overview of how regular expressions and context-free grammars/expressions are essentially sub-languages of our type system.}

\section{Semantics}
\label{sec:semantics}
\subsection{Semantics in Grammars}
\label{subsec:seming}
As mentioned earlier, the interpretation of grammars as functions from
$\String$ to $\Set$ serves as our intended semantics. When designing the syntax of this formalism, we were often
informed by this model. That is, there is an interpretation
$\llbracket \cdot \rrbracket$ of our syntax that provides grammar-based
semantics.

\begin{enumerate}
  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$
  \item $\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
        \begin{itemize}
          \item The semantics of the linear $\Pi$-type is
                indeed a dependent function out of the
                semantics of $X$.
        \end{itemize}
  \item $\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$
        \begin{itemize}
          \item The semantics of the linear $\Sigma$-type is
                a dependent pair.
        \end{itemize}
  \item $\sem{I} \gamma w = \{ () \pipe w = \epsilon \}$
        \begin{itemize}
          \item The only string accepted by $I$ is the empty string.
        \end{itemize}
  \item
        $\sem{A \otimes B} \gamma w = \Sigma(w1,w2:\Sigma^*) \left( (w_{1}w_{2} = w) \wedge \sem{A} \gamma w_1 \times \sem{B} \gamma w_2 \right)$
        \begin{itemize}
          \item The string $w$ matches $A \otimes B$
                precisely when it may be split into two
                pieces such that the left one matches $A$
                and the right one matches $B$.
        \end{itemize}
  \item
        $\sem{A \lto B} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (w_aw) \right)$
        \begin{itemize}
          \item The semantics of the linear function type is
                a dependent function that turns $A$-parses
                of a substring into $B$-parses of an
                extended string.
        \end{itemize}
  \item
        $\sem{B \tol A} \gamma w = \Pi(w_a:\Sigma^*) \left( A \gamma w_a \Rightarrow B\gamma (ww_a) \right)$
  \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$
        \begin{itemize}
          \item The semantics of the least fixed point
                operator is indeed the least fixed point of sets.
        \end{itemize}
\end{enumerate}

\subsection{Canonicity}

@@

\subsection{Semantics in Languages}
% Recall from \cref{sec:synindsyn} that every formal grammar induces a formal language. This corresponds to another
interpretation $L(\cdot)$ of our syntax. In particular, the semantics of a
linear type $A : L_{i}$
from \cref{subsec:seming} are augmented such that

\[
  L(A) = \{ w \in \String : \sem A w \text{ is nonempty} \}
\]

\subsubsection{Weak Equivalence}
Recall that an isomorphism of grammars defined a strong equivalence. We may
additionally define a \emph{weak equivalence} of grammars when two grammars
define the same language.

That is, $A$ and $B$ are weakly equivalent when $L(A) = L(B)$. Note that this is
a semantic notion, but this just as well captured syntactically by the existence
of two judgments $\Gamma ; x : A \vdash M : B$ and
$\Gamma ; x : B \vdash N : B$, which are not necessarily mutually inverse.

>>>>

Omission of the structural rules of a deductive system, such as in
linear logic \cite{GIRARD19871}, offers precise control over how a value is used
in a derivation. Namely, linear logic omits the weakening and contraction rules
to ensure that every value in context exactly once. This control enables
\emph{resource-sensitive} reasoning, where we may treat a resource as
\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
as we may treat characters of a string finite resources that are consumed at
parse-time. That is, once a particular substring has been parsed, a parser need not go
over it again. One may then envision a linear type system where the types comprise
formal grammars generated over some alphabet $\Sigma$, and the type constructors
correspond precisely to inductive constructions on grammars --- such as
conjunction, disjunction, concatenation, etc.

@@

\steven{Maybe there is more to say here, or perhaps later, that the only thing
  to gain parsing-wise from a non-dependent non-linear fragment are the things that are
  definiable via $G$. Except that these are not interesting, as they are
  necessarily empty}

@@

\subsection{Syntax}
\label{subsec:syntax}

We instantiate an LNL theory with dependent types as a deductive
setting to axiomatize the structure of a Chomsky category.

The structrual judgements are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
\cref{fig:inttyping}. These are all just as they appear in
\cite{krishnaswami_integrating_2015}.

@@

    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \\
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgmental equality}
  \label{fig:jdgeq}
\end{figure}

\subsection{Formal Grammars as Linear Types}
The linear typing judgment in our syntax takes on the following schematic form
$\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
our syntax. The intended semantics of these linear types are formal grammars.
That is, the linear typing system is designed to syntactically reflect the
behavior of formal grammars. For this reason, we may often interchangeably use
the terms ``linear type'' and ``grammar''.

The term $a$ is an inhabitant of type $A$, which is thought of as a parse tree of
the grammar $A$. The core idea of this entire paper follows precisely from this
single correspondence: grammars are types, and the inhabitants of these types
are parse trees for the grammars.

$\Gamma$ reprsents a non-linear context, while $\Delta$ represents a
\emph{linear context} dependent on $\Gamma$. These linear contexts behave
substructurally. As stated earlier, they do are linear --- they do not obey
weakening or contraction --- because a character is exhausted once read by a
parsing procedure. Moreover, the characters in strings appear in the order in
which we read them. We do not have the freedom to freely permute characters,
therefore any type theory that is used to reason about formal grammars ought to
omit the structural rule of exchange as well. This means that every variable
within $\Delta$ must be used \emph{exactly once} and \emph{in order of
  occurrence}. Thus, we can think of the linear contexts as an ordered list of
limited resources. Once a resource is consumed, we cannot make reference to it
again. Variables in a linear context then act like building
blocks for constructing patterns over strings.

We give the base types and type constructors for linear terms. As the
interpretation of types as grammars in $\Set^{\String}$ serves as our intended
semantics, we simultaneously give the interpreation $\sem \cdot$ of the
semantics as grammars.

\paragraph{Linear Unit}
First, the linear unit $I$ may be built in the empty linear
context\footnote{When appropriate, statements like ``the empty linear context''
  may be taken polymorphically over all non-linear contexts that they might
  depend on.} $\cdot \dashv I$. $I$ serves as the unit for the operation
$\otimes$ described below.

As a grammar,

\[
  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
\]

That is, $I$ maps strings to the proposition that they are the empty string. $I$
is clearly only inhabited by the empty string $\varepsilon$, for which the
outputted set contains a single parse tree.

\paragraph{Base Types}
For each character $c$ in the alphabet $\Sigma$, we include a base type at the
lowest universe level $c : L_{0}$.

The grammar interpretation for characters is quite similar to that of $I$.
\[
  \sem {c} \gamma w = \{ () ~|~ w = c \}
\]

The grammar for the character $c$ likewise maps strings $w$ to the proposition that
they are equal to $c$.

\paragraph{Tensor Product}
The tensor product of two linear types is the first place where the ordering on
contexts really takes effect. That is, the tensor product of two types is
non-commutative.
When context $\Delta$ forms a linear term of
type $A$, and $\Delta'$ forms a linear term of type $B$, then the context
extension $\Delta, \Delta'$ forms a linear term of type $A \otimes B$.

As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
linear types $A$, the following equalities hold:

\[ I \otimes A \equiv A \otimes I \equiv A \]

In the grammar semantics,

\[
  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
\]

The string $w$ matches $A \otimes B$
precisely when it may be split into two
pieces such that the left one matches $A$
and the right one matches $B$.

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types.

\steven{Say more about the adjunctions}

Further, for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application/modus ponens.

\[
\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
\]

The interpretation of linear function types at a string $w$ is a linear function
that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
string $w_{a} w$.

\[
  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
\]

That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types the same notion of derivative extends to the grammars of our type theory.

Of course, all the above statements for the left function type also have
corresponding analogues for the
right-handed counterpart.

\paragraph{LNL Dependent Types}
Given a non-linear type $X$, we may form both the dependent product type
$\LinPiTy x X A$ and the dependent pair type $\LinSigTy x X A$ as linear types
where $x$ is free in $A$.

$\sem{\LinPiTy x X A} \gamma w = \Pi(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$

The grammar semantics of the linear product type is
indeed a dependent function out of the
semantics of $X$.

$\sem{\LinSigTy x X A} \gamma w = \Sigma(x:\sem{X}\gamma) \sem{A}(\gamma,x) w$

Likewise, the grammar semantics of the linear dependent pair type is a dependent
pair in $\Set$.

\steven{TODO example of sum as an admissible usage of $\LinSigTy b \Bool A$}

\paragraph{Universal Type}
The universal type $\top$ may be formed in any context.

\[ \sem{\top} \gamma w = \{ \ast \}\]

Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.

\paragraph{Empty Type}
The empty type $\bot$ has no inhabitants.

\[ \sem {\bot} \gamma w = \emptyset \]
\steven{TODO need to include a case for $\bot$ in the inference rules}

\paragraph{Pair Type}
Given a term of type $A$ in context $\Delta$ and a term of type $B$ \emph{also
  in context} $\Delta$, we may form a pair in context $\Delta$ of type
$A \ampersand B$.

\[ \sem{A \amp B }\gamma w = \sem{A} \gamma w \times \sem {B} \gamma w \]

The grammar semantics of $A \amp B$ denotes a pair of a parse of $A$ and a
parse of $B$ \emph{of the same string}. In this sense, $\amp$ acts a
conjunction between grammars --- meaning a string only matches $A \amp B$
if it simlutaneously matches each of $A$ and $B$.

\paragraph{The $G^{-1}$ Modality}
Turn nonlinear into linear

\steven{Call this $F$ or $G^{-1}$?}

\steven{Why can't these map any nonlinear type into linear ones? instead of just
those that are already linear types lifted by $G$?}

\steven{TODO G inverse}
  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$

\paragraph{Recursive Types}
isorecursive types
\steven{TODO recursive types}
  % \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$

@@

\subsection{Subtheories}
\paragraph{Regular Expressions} Restricting to regular expressions with the
obvious connectives forms a subtheory of our type theory. Perhaps there is
something to say here about how this reduces to Kleene algebra performed inside
of a Kleene cat?

\steven{TODO regular expressions}

\paragraph{$\mu$-Regular Expressions} Regex + fixed point equivalent to context
free \cite{leiss}.

\steven{TODO $\mu$ regular expressions}

\steven{Every after here has not been touched}

\section{Canonicity}

@@
